### 2021-04-21 final-report.latex/intro.tex
fix Chris' notes in Methods
beef up intro

Quantum field theory (QFT) is an extraordinarily successful framework which describes a range of physical phenomena. Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics. In condensed matter physics, effective field theories model emergent effects such as phonons and quasiparticles. Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eight significant figures.\cite{weisskopf1981}

However, this power comes at a cost: the study of quantum fields is rife with infinities. A na\"ive treatment of quantum field theory produces divergent values for physical quantities, a clearly impossible result. Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the so-called \textit{renormalization group}. However, the technique fails with perturbatively nonrenormalizable theories.

One such example is the \textit{non-linear sigma model}, a prototypical theory in both condensed matter and particle physics. In solid-state systems, this model describes Heisenberg ferromagnets and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), exhibiting characteristic features such as a mass gap and asymptotic freedom. \citeneeded

In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions one dimension of space, one dimension of time). This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action.

Since the non-linear sigma model cannot be renormalized perturbatively, we cannot study these topological effects with normal perturbative techniques. An alternative solution is placing the field on a discretized lattice, a technique originally used for quantum chromodynamics. In this scenario, field configurations become computationally calculable. This process introduces an nonphysical length scale $a$, the lattice spacing. Therefore, we expect any physical result to converge in the continuum limit, i.e. when $a\rightarrow 0$. However, this is not always the case as observables mix on the lattice, leading to divergences. As an example, states of definite angular momentum mix when discretized, a clear violation of the angular momentum commutation relations.

The gradient flow is a technique designed to remove these divergences. By dampening high-momentum fluctuations, the gradient flow reduces power-divergent mixing and make observables finite on the lattice.\cite{monahan2016} In quantum chromodynamics, previous studies have verified the ability of the gradient flow to make observables finite\citeneeded. Due to its success in QCD, there has been interest in using the gradient flow to finitize the topological susceptibility in the 1+1 $O(3)$ non-linear sigma model.\cite{bietenholz2018}.

%goals here

@@

@@

%\section{Main Experimental Goal}

>>>>

Quantum field theory (QFT) is an extraordinarily successful framework which describes a range of physical phenomena. Paired with the Standard Model, QFT provides the prevailing basis for all small-scale physics (that is, where general relativity does not apply) and is the fundamental tool for studying particle physics. QFT is also plays an critical role in condensed matter physics through effective field theories which model emergent effects such as phonons and quasiparticles. Compared to experiment, QFT is remarkably accurate, famously matching the experimental value for the electron $g$-factor to eleven significant figures \cite{odom2006}, arguably the most accurate prediction in all of science.

However this power comes at a cost: the study of quantum fields is rife with infinities. A na\"ive treatment of quantum field theory produces divergent values for physical quantities, a clearly impossible result. Since the 1950s, this issue has been resolved for a large number of models--- most notably quantum electrodynamics--- through perturbation theory and the renormalization group which exploits the freedom of parameters such as mass and electric charge. Since these values cannot be directly measured, renormalization allows them to become infinite, thereby cancelling the infinities in physical results. However, there exists a set of theories which as not perturbatively renormalizable.

One such example is the \textit{non-linear sigma model} (NLSM), a prototypical theory in both condensed matter and particle physics. In solid-state systems, this model describes Heisenberg ferromagnets \cite{callan1985} and in nuclear physics, it acts as a prototype for quantum chromodynamics (QCD), the gauge theory that describes the strong nuclear force. In general, the NLSM shares key features with non-Abelian gauge theories (such as QCD) like a mass gap and asymptotic freedom \cite{polyakov1975} and therefore is a useful model for exploring these phenomena in simpler systems.

In this study, we specifically consider the $O(3)$ non-linear sigma model in $1+1$ dimensions (one dimension of space, one dimension of time). This theory exhibits topological properties such as \textit{instantons}, or classical field solutions at local minima of the action in Euclidean space. Solutions such as these are ``topologically protected'', meaning they cannot evolve into the vacuum state via small fluctuations. Therefore, it is critically important to take topology into account when considering quantum field theories in a cosmology and high energy physics \cite{goddard1986}. Additionally, topological stability may become a key tool for fault-tolerant quantum computers \cite{kitaev1997}. In these devices, topology could protect the delicate quantum states necessary for information processing.

Since the NLSM in not perturbatively renormalizable, we require nonperturbative techniques to study topological effects. An alternative solution is to place the field on a discretized Euclidean lattice, a technique originally used for quantum chromodynamics \cite{wilson1974}. After this transformation, fields become computationally calculable. This process introduces the lattice spacing as a length scale $a$. 
% something here
Therefore, we expect any physical result to converge in the continuum limit, i.e. when $a\rightarrow 0$. However, this is not always the case as observables mix on the lattice, leading to divergences. As an example, states of definite angular momentum mix when discretized, a clear violation of the angular momentum commutation relations.

In this study, we focus on one observable that diverges in the continuum: the topological susceptibility. The convergence of the topological susceptibility in the 1+1 $O(3)$ NLSM has been the subject of debate for the past four decades \cite{berg1981,bietenholz2018}. In QCD, mathematical techniques have proven that this quantity vanishes in the continuum limit \cite{giusti2004}, however a similar line of reasoning for the NLSM has not been successful \cite{bietenholz2018}.

To remedy this issue, we consider the gradient flow, a technique designed to remove divergences in operators. By dampening high-frequency fluctuations, the gradient flow reduces terms which scale with the inverse lattice spacing, making some observables finite on the lattice \cite{monahan2016}. In QCD, the gradient flow successfully makes the topological susceptibility finite in numerical calculations \cite{bruno2014}, corroborating the analytical result in \cite{giusti2004}. This success has motivated the usage of the gradient flow to calculate the topological susceptibility in the 1+1 $O(3)$ NLSM, though the observable still diverges in the continuum limit \cite{bietenholz2018}.

A second perspective on the topological susceptibility arises from the introduction of a $\theta$ term into the field Lagrangian. This term drives the vacuum state into a topological phase \cite{alles2007}. Differentiating the field's partition function with respect to $\theta$ yields a value proportional to the topological susceptibility. The effect of nonzero $\theta$ on the theory therefore should reflect the divergence in the continuum limit.

In this paper we verify the result the divergence of the topological susceptibility and develop a clearer picture of how the $\theta$-term affects the topology of the NLSM.

@@

\section{Summer Research}

A portion of this work began during the summer of 2021 using funding from the 1693 Scholars Program. This preliminary research consisted of literature review and numerical tests with an Ising model simulation as well as an initial implementation of the $\phi^4$ model. The $\phi^4$ calculation performed in this study and the entirety of the NLSM portion occurred during the academic year as part of the PHYS 495/496 Honor course.

@@
